---
title: "Most Impactful Calamities in US Health & Economics"
author: "Nikolaos Perdikis"
date: "18 6 2019"
output: 
  html_document: 
    keep_md: yes
---
## Synopsis
This report examines a set of data from the NOAA storm database. Event types are groupped to display trends, and the most harmful events in terms of life and property are displayed in graphics and tabular data reports. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Processing
```{r ReadBZ2file, cache=TRUE}
setwd("D:/Github/Nikolas")
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileURL,"stormdata.bz2")
NOAADB <- read.csv("stormdata.bz2",stringsAsFactors = FALSE)
```

The data source is an excerpt from the [NOAA Storm Database](https://www.ncdc.noaa.gov/stormevents/). It will be mentioned in the rest of the document as "the database"

### Validation of the Data - Latitude/Longitude and US state
The code below was used to identify valid US state codes (Continental 50 + Hawaii), since there is a much larger number of unique entries for 'STATE' in the database.  
**The state codes below do not point to a valid state name:**
```{r}
#Function below will return all non-valid state code entries
StateName <- function(a) {
    b <- state.name[match(a,state.abb)]
    ifelse (is.na(b),a,b)
    }
res <- subset(NOAADB,StateName(NOAADB$STATE)==NOAADB$STATE)
unique(res$STATE)
```

**Geo Coordinates**
There are events with no LAT/LON coordinates, but worse the LONGITUDE contains a single negative value which resolves to state wth code "GU" and a number of positive values. In order for a map plot to take place it would make sense to take the following assumptions:  
LATITUDE: Needs to be divided by 100 to provide a correct value, and further assumption that in that case the format will be decimal degrees (and not degrees/minutes)  
LONGITUDE: All values are to assumed to be multiplied by -1 to become negative and divided by 100 to provide decimal degrees format (as with LATITUDE) 



```{r FindDifferentEVTYPE}
##str(NOAADB)
n_evt <- nrow(NOAADB)
n_ev_type <- unique(NOAADB$EVTYPE)
count_type <- length(n_ev_type)
```

The data provided contains **`r n_evt` events**, categorised into **`r count_type` levels**. Those are the different types of calamities encountered. The data source also contains such information as Begin and End Date/Time/Location, number of Fatalities, number of Injuries and the amount of Damage in Property and Crops.  

In this research, three different findings will be presented:  

1) The top ten types of incident that has been more harmful, in terms of both human life and financial damages
2) The top ten isolated incidents that has caused the most harm, in terms of human life and financial damages
3) A detailed review of a specific event (Heat)

To assess the most harmful events in terms of human life, a data frame was created, groupping the data source by event type, and summing the equivalent fields of Property and Crops damage in the database. 

```{r LoadLibraries, message=FALSE}
library (dplyr)
library(ggplot2)
library(lubridate)
```
Conversion of Date fields:  
We need to convert the date fields from factors to date format, so we can perform comparisons and filtering based on ranges. First we need to only keep the date part, as each entry has a trailing hours value that is always set at midnight.

```{r ConvertDatesAddYear, cache=TRUE}
#Extract the time entry, keep just month/day/year
NOAADB$BGN_DATE <- sub(" .*","",NOAADB$BGN_DATE)
NOAADB$END_DATE <- sub(" .*","",NOAADB$END_DATE)
NOAADB$BGN_DATE <- as.Date(NOAADB$BGN_DATE,"%m/%d/%Y")
NOAADB$END_DATE <- as.Date(NOAADB$END_DATE,"%m/%d/%Y")
#Display earliest recorded incident
min(NOAADB$BGN_DATE)
#Store the year as a different column in our dataset
NOAADB$YEARS <- year(ymd(NOAADB$BGN_DATE))
#Gather some basic information, such as earliest/latest recording of an event, its total occurences, maximum cost in human life and economics
```
Having fixed the date fields, next step is to gather some basic information, such as earliest/latest recording of an event, its total occurences, maximum cost in human life and economics

```{r FirstGroupping}
inf <- NOAADB %>% 
    group_by(EVTYPE) %>% 
    summarize(number=n(),max_fatal=max(FATALITIES),tot_fatal=sum(FATALITIES),min(BGN_DATE),max(BGN_DATE))
# Order by Most Lethal Incident
inf[order(-inf[3]),]
#And display by number of total fatalities
inf[order(-inf[4]),]
```

### Key Early Findings
1) Groupping the Event Type by categories of similar incidents would yield more meaningful numbers.
2) While some categories have many decades of data (TORNADO, TSTM WIND), others have not have records until almost 4 decades later. So attempting a comparison of the human and financial damages between such categories may not provide realistic information.

### Data Processing Continued - Groupping Events into Categories  

```{r GrouppedEVTYPE}
# Convert factor to char to avoid error message "invalid factor level, NA generated"
NOAADB$EVTYPE <- as.character(NOAADB$EVTYPE)

#Define the group patterns
category_HEAT <- c("HEAT","WARMTH","HIGH TEMPERATURE")
category_COLD <- c("COLD","LOW TEMPERATURE","WINTER")
category_FLOODS <- c("FLOOD","CURRENT","SURF","STREAM")
category_RAIN_SNOW <- c("RAIN","SNOW","STORM","BLIZZARD")
category_WIND <- c("WIND","HURRICANE")
category_FIRE <- c("FIRE")
category_TORNADO <- c("TORNADO","FUNNEL")

#Collect all values
matches <- unique(grep(paste(category_HEAT,collapse="|"),NOAADB$EVTYPE,value=TRUE))
#For example, for the "HEAT" category, all the different event types below will be groupped into one:  
matches
#Replace with a single value in the group                  
NOAADB$EVTYPE[NOAADB$EVTYPE %in% matches] <- "HEAT"

#Repeat for all groups
matches <- unique(grep(paste(category_COLD,collapse="|"),NOAADB$EVTYPE,value=TRUE))
NOAADB$EVTYPE[NOAADB$EVTYPE %in% matches] <- "COLD"

matches <- unique(grep(paste(category_FLOODS,collapse="|"),NOAADB$EVTYPE,value=TRUE))
NOAADB$EVTYPE[NOAADB$EVTYPE %in% matches] <- "FLOODS"

matches <- unique(grep(paste(category_RAIN_SNOW,collapse="|"),NOAADB$EVTYPE,value=TRUE))
NOAADB$EVTYPE[NOAADB$EVTYPE %in% matches] <- "RAIN / SNOW"

matches <- unique(grep(paste(category_WIND,collapse="|"),NOAADB$EVTYPE,value=TRUE))
NOAADB$EVTYPE[NOAADB$EVTYPE %in% matches] <- "WIND"

matches <- unique(grep(paste(category_FIRE,collapse="|"),NOAADB$EVTYPE,value=TRUE))
NOAADB$EVTYPE[NOAADB$EVTYPE %in% matches] <- "FIRE"

matches <- unique(grep(paste(category_TORNADO,collapse="|"),NOAADB$EVTYPE,value=TRUE))
NOAADB$EVTYPE[NOAADB$EVTYPE %in% matches] <- "TORNADO"

# This also works: filter(stdata,grepl(paste(cat_HEAT,collapse = "|"),EVTYPE)) but how to store back to the df
# Next, prepare for two tables, most lethal incident and incident type with most fatalities
inf <- NOAADB %>% 
    group_by(EVTYPE) %>% 
    summarize(INCIDENTS=n(),FATALITIES_BY_INCIDENT=max(FATALITIES),TOTAL_FATALITIES=sum(FATALITIES))
```

### Most Lethal Incident, based on number of Fatalities and its Type
```{r}
# Order by Most Lethal Incident
inf[order(-inf[3]),]
```

### Incident Group responsible for the biggest number of Fatalities
```{r}
#And display by number of total fatalities
inf[order(-inf[4]),]
```

From the numbers above, it can be seen that:  
1) TORNADO has the highest number of fatalities in total, but the recorded dates of such events far exceed the dates of others. There are tornadoes recorded as far back as 1950, while events related to high temperature have not been recorded before 1993.  
2) **HEAT comes out with the most lethal event (583 fatalities)** in the entire database, second highest total cause of loss of life in the database, and this, even with reocords appear after 1993.

### Plot for the Top 10 fatality causes
```{r OrderPlotFatalities}
res_all <- NOAADB %>% group_by(EVTYPE) %>%
    summarise(Fatalities=sum(FATALITIES))
#order by fatalities
res_all_ordered <- res_all[order(-res_all[2]),]
#keep Top 10
res_all_ordered <- res_all_ordered[1:10,]


ggplot(aes(x=reorder(EVTYPE,Fatalities),y=Fatalities),data=res_all_ordered) +
    geom_point()+
    scale_y_continuous(breaks=0) +
    labs(title="Fatalities per Event Type Across the US",x="Type of Event",y="Total Fatalities")+
    geom_text(aes(label=Fatalities),color="black",size=3,vjust=-0.5)+
    theme(axis.text.x=element_text(angle=90))

```
  
### Top 10 most Expensive Event Types in terms of Property and Crops Loss

```{r}
res_all <- NOAADB %>% group_by(EVTYPE) %>%
    summarise(Damages=sum(PROPDMG+CROPDMG))
#order by Damages
res_all_ordered <- res_all[order(-res_all[2]),]
#keep Top 10
res_all_ordered <- res_all_ordered[1:10,]


ggplot(aes(x=reorder(EVTYPE,Damages),y=Damages),data=res_all_ordered) +
    geom_bar(stat="identity",fill = "forestgreen")+
    scale_y_continuous(breaks=0) +
    labs(title="Damages (USD) per Event Type Across the US",x="Type of Event",y="Total Damages")+
    geom_text(aes(label=Damages),color="black",size=3,vjust=-0.5)+
    theme(axis.text.x=element_text(angle=90))
```
  
### Geographical distribution of Events
In this section, we group events ordered by the frequency of different US states they appear in. Events impacting the largest number of states are on top.  

```{r}
res <- subset(NOAADB,StateName(NOAADB$STATE)!=NOAADB$STATE)
res <- res %>% group_by(EVTYPE) %>% summarise(n=length(unique(STATE)))
res[order(-res[2]),]
```

### The case of Heat

The following plot displays the HEAT related incidents through the years. What clearly stands out is the 1995 peak from Heat related deaths. This is a known and documented event, and more information can be found [in this Wikipedia Article](1995_Chicago_heat_wave). Also another "positive" trend is that while the number of heat incidents rises (as anticipated through the Global Warming phenomenon) the number of casualities from heat decreases.

```{R PlotHeat}
heat <- subset(NOAADB,EVTYPE=="HEAT")
heat_year <- heat %>%
    group_by(YEARS) %>% summarize(Incidents=n(),Fatalities=sum(FATALITIES))
#The data
heat_year
#And the plot
plot(heat_year)
```

## Results
All of the world,environment and the climate have undergone drastic changes from the early dates of this report until its latest. Technology has certainly helped avoid some types of natural disasters and help against climate extremes, but itself, with its large energy footprint may also be attributed as a main contributor to said environmental changes. While the database covers far too many decades to be able to draw a singular conclusion, it can be seen that even a large geographical area as the Continental US has common issues among its many regions. Numbers that were a surprise were how Heat far outnumbers Cold in terms of fatalities and Injuries, and how Heat in total is a very large contributor to extreme events. 